Global patterns and drivers of vertebrate verticality

### 2024-09-05 verticality_future_mkd.Rmd
intro revisions

Vertical stacking of species is an emblematic signature of biologically rich regions and is associated with patterns of species richness from local to global scales [@macarthurPopulationEcologyWarblers1958; @macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019; @bashamVerticalStratificationPatterns2023a]. At local scales, species assemblages may be stratified across vertical resource and climate gradients from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023]. Tall, multi-layered forests increase both habitat volume and niche opportunities within the habitat, acting effectively as multi-floored apartments that enable species to vertically partition resources [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. By reducing competition, vertical stratification may be a mechanistic link to positive relationships between canopy height and species richness [@culbertInfluenceVerticalHorizontal2013; @fengForestCanopyHeight2020; @gillespieDistributionBirdSpecies2001; @oliveiraVerticalStratificationInfluences2019; @wagnerRemotelySensedTree2023]. However, canopy height does not consistently impact species richness at regional to global scales, with notable declines in the strength of the relationship at higher latitudes where the diversity of arboreal species generally declines [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @oliveiraVerticalStratificationInfluences2019]. Examining the impacts of environmental drivers of vertical stratification may thus add nuance to our understanding of the impact of canopy height on species richness.

At regional to global scales, climate may constrain vertical stratification in some regions despite the presence of tall forests. Vertical climate gradients are characterized by fossorial and terrestrial habitats that exhibit lower thermal variability and greater moisture availability than arboreal habitats [@scheffersTropicalMountainPasses2018; @leahyVerticalNicheElevation2021; @defrenneGlobalBufferingTemperatures2019]. At high latitudes, low minimum temperatures in the canopy may exceed physiological thresholds and constrain species to lower vertical positions [@oliveiraVerticalStratificationInfluences2019; @bakenSalamanderArborealityLimited2021]. In wet tropical forests, physiological constraints decline, which may allow species to stratify across the entire vertical habitat gradient and contribute to the high diversity of arboreal species found in these habitats [@oliveiraVerticalStratificationInfluences2019]. We therefore expect the impact of canopy height on verticality to vary across large spatial extents, exhibiting a positive impact where climate supports arboreal life strategies and a neutral impact where climatic conditions pose greater physiological challenges.

Spatial variation may also occur in the relationship between species richness and  verticality [@oliveiraVerticalStratificationInfluences2019]. Though prior studies have shown positive relationships between verticality and richness, the strength of this relationship is variable and generally more positive in tropical regions [@cabralFunctionalTraitsPhylogeny2022; @oliveiraVerticalStratificationInfluences2019]. Where climatic conditions allow canopy height to positively impact verticality, as may be expected in the wet tropics, the positive relationship between verticality and richness may be stronger because the addition of a vegetation layer allows the addition of arboreal species to the assemblage. In contrast, where taller canopies do not facilitate higher verticality, the relationship between verticality and richness may be weak.

As the climate changes, constraints on arboreal life strategies may shift with potential impacts on assemblage verticality and its association with species richness. In tropical lowlands, increasing temperatures and declining dry season precipitation (*cite*) may pose physiological challenges for arboreal animals and cause them to disperse toward higher elevations, lower vertical positions, or become locally extirpated [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019]. These scenarios would reduce assemblage verticality in the lowlands, reflecting climate-induced shifts in assemblage thermal tolerance [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024]. At high latitudes, increasing minimum temperatures could reduce climatic constraints on arboreal species and facilitate range shifts that increase verticality.

@@

Despite widespread agreement that verticality influences biodiversity patterns, global patterns of verticality, its environmental drivers, its relationship with species richness, and its vulnerability to climate change have not been comprehensively described for vertebrates at a global scale. To improve our understanding of the nuances of the relationships between climate, vegetation structure, verticality, and species richness, we integrate range maps of `r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians with climate and vegetation data to examine global patterns and environmental drivers of vertebrate verticality in wooded biomes. We then examine whether the relationship between verticality and canopy height mediates the relationship between verticality and species richness. Finally, we predict how climate change may impact global patterns of verticality.

>>>>

In forest ecosystems, species assemblages may be stratified across vertical resource and climate gradients from beneath the ground to the top of the canopy, i.e., vertical stratification [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023]. Tall, multi-layered forests increase both habitat volume and niche opportunities, enabling species to partition across vertical strata akin to multi-floored apartments [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. This vertical stacking of species has become an emblematic signature of biologically rich regions and is associated with patterns of species richness from local to global scales [@macarthurPopulationEcologyWarblers1958; @macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019; @bashamVerticalStratificationPatterns2023a]. As such, vertical stratification may be a mechanistic link to positive relationships between canopy height and species richness [@culbertInfluenceVerticalHorizontal2013; @fengForestCanopyHeight2020; @gillespieDistributionBirdSpecies2001; @oliveiraVerticalStratificationInfluences2019; @wagnerRemotelySensedTree2023]. However, canopy height does not consistently impact species richness at regional to global scales, with notable declines in the strength of the relationship at poleward latitudes where the diversity of arboreal species generally declines [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @oliveiraVerticalStratificationInfluences2019].

Tall forests contain steep vertical climate gradients characterized by lower thermal variability and greater moisture toward the ground [@scheffersTropicalMountainPasses2018; @leahyVerticalNicheElevation2021; @defrenneGlobalBufferingTemperatures2019]. At temperate and boreal latitudes, low temperature and moisture in the canopy may exceed physiological thresholds and constrain species to lower vertical positions [@oliveiraVerticalStratificationInfluences2019; @bakenSalamanderArborealityLimited2021]. In wet tropical forests, physiological constraints decline, which may allow species to stratify across the entire vertical habitat gradient [@oliveiraVerticalStratificationInfluences2019]. We therefore expect a globally varying relationship between canopy height and verticality pending baseline regional conditions from positive in tropical to neutral in temperate and boreal systems.

Spatial variation may also occur in the relationship between verticality and species richness [@oliveiraVerticalStratificationInfluences2019]. Though prior studies have shown positive relationships between verticality and richness for select taxa, the strength of this relationship is variable and generally more positive in tropical regions where climatic conditions pose fewer physiological constraints on arboreal strategies [@cabralFunctionalTraitsPhylogeny2022; @oliveiraVerticalStratificationInfluences2019]. Niche expansion, characterized by vertical partitioning of the foerst, may allow this pattern to occur. In contrast, the relationship between verticality and richness may be weak where taller canopies do not facilitate higher verticality.

As the climate changes, constraints on arboreal life strategies may shift with potential impacts on assemblage verticality and its association with species richness. In tropical lowlands, increasing temperatures and declining dry season precipitation (*cite*) may pose physiological challenges for arboreal animals and cause them to disperse toward higher elevations, lower vertical positions, or become locally extirpated [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019]. These scenarios would reduce assemblage verticality in the lowlands, akin to studies that show shifts in ecological processes such as climate-induced shifts in assemblage thermal tolerance [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024]. At poleward latitudes, increasing minimum temperatures could reduce climatic constraints on arboreal species and facilitate range shifts that increase verticality.

@@

Despite decatdes of ecological inquiry into the relationship between forest height and species richness, little is known as to whether tall forests ubiquitously support niche filling and thus promote greater species coexistence per unit of area (high alpha or gamma diversity) To improve our understanding of the nuances of the relationships between climate, vegetation structure, verticality, and species richness, we integrate range maps of vertebrates (`r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians) with climate and vegetation data to examine global patterns and drivers of verticality in wooded biomes. We then examine whether the dynamic of filling vertical niche space and its effects of species richness is informed by the relationship between verticality, canopy height, and climate. Finally, we predict how climate change may impact global patterns of verticality.
